---
title: "SBBG Elings Park Analysis"
author: "Caroline Shepherd"
date: "2024-02-27"
output: 
  html_document:
    theme: cerulean
    code_folding: hide
    highlight: monochrome
    toc: yes
    toc_depth: 5
    toc_float: yes
    collapsed: yes
    smooth_scroll: yes
    number_sections: no
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
# read in packages

library(janitor)
library(tidyverse)
library(here)
library(dplyr)

```

## Introduction:

This code aims to synthesize and visualize data collected on Biomass (kg/m2) and Mortality of plants in the Elings Park Restoration Project. The Santa Barbara Botanic Garden, in partnership with Channel Islands Restoration, is transforming a 1-acre plot of land previously full of invasive weeds, into a better-functioning native habitat.

In this analysis, we explore the relationship between different weed treatments (sheet mulching and plastic tarping) and plant mortality and growth. 

```{r}
# read in the data
# Make sure you have the most updated data
elings_data <- read.csv(here("ElingsMonitoringData25March2024.csv")) %>% clean_names() %>% 
  rename(width1_cm = width1_cm_widest_diameter, # changing column names to be less annoying
         width2_cm = width2_cm_perpendicular_to_w1_measurement) %>% 
  mutate(width1_m = width1_cm/100,  #making a width column in meters
         width2_m = width2_cm/100) %>% 
  mutate(veg_type = case_when(species_id == "BOBA" ~ "herb",
                              species_id == "STPU" ~ "herb",
                              species_id == "BRCA" ~ "herb",
                              species_id == "CAMA" ~ "herb",
                              species_id == "COFI" ~ "herb",
                              species_id == "EPCA" ~ "herb",
                              species_id == "GRCA" ~ "herb",
                              species_id == "LECO" ~ "herb",
                              species_id == "SODO" ~ "herb",
                              species_id == "SPAM" ~ "herb",
                              species_id == "ACGL" ~ "shrub",
                              species_id == "AGSH" ~ "shrub",
                              species_id == "ARCA" ~ "shrub",
                              species_id == "ATLE" ~ "shrub",
                              species_id == "BAPI" ~ "shrub",
                              species_id == "BEEM" ~ "shrub",
                              species_id == "ENCA" ~ "shrub",
                              species_id == "ERFA" ~ "shrub",
                              species_id == "EUMI" ~ "shrub",
                              species_id == "HEWH" ~ "shrub",
                              species_id == "ISME" ~ "shrub",
                              species_id == "LEGI" ~ "shrub",
                              species_id == "LYCA" ~ "shrub",
                              species_id == "MAFA" ~ "shrub",
                              species_id == "PEAR" ~ "shrub",
                              species_id == "RHCR" ~ "shrub",
                              species_id == "RHIN" ~ "shrub",
                              species_id == "SAAP" ~ "shrub"))

# need to create function to calculate biomass


# group by plot and sum biomass 

biomass_plot <- elings_data %>% 
  group_by(plot_id) %>% 
  summarise(biomass_sum = sum(biomass_kg_m2))

# group by plot and sum mortality
# I double checked and this is correctly summing # of alive plants

mort_plot <- elings_data %>% 
  group_by(plot_id) %>% 
  summarise(mort_sum = sum(mort_status, na.rm = TRUE))

# group by treatment and sum biomass

biomass_treatment <- elings_data %>% 
  group_by(treatment) %>% 
  summarise(biomass_sum = sum(biomass_kg_m2))

# group by treatment and sum mortality

mort_treatment <- elings_data %>% 
  group_by(treatment) %>% 
  summarise(mort_sum = sum(mort_status, na.rm = TRUE))

```


### Data exploration

Looking at trends in mortality and biomass based on plot and treatment.
```{r}

# explore the data! look at trends!

#graphing biomass by plot
ggplot(data = biomass_plot, aes(x = plot_id, y = biomass_sum)) +
  geom_col() +
  theme_bw() +
  labs(x = "Plot ID",
       y = "Total Biomass (kg/m2)")

# graphing biomass by treatment
ggplot(data = biomass_treatment, aes(x = treatment, y = biomass_sum)) +
  geom_col() +
  theme_bw() +
  labs(x = "Treatment",
       y = "Total Biomass (kg/m2)")

# graphing mortality by plot
ggplot(mort_plot, aes(x = plot_id, y = mort_sum)) +
  geom_col() +
  theme_bw() +
  labs(x = "Plot ID",
       y = "Total Number of Live Plants")

# graphing mortality by treatment
ggplot(data = mort_treatment, aes(x = treatment, y = mort_sum)) +
  geom_col() +
  theme_bw() +
  labs(x = "Treatment",
       y = "Total Number of Live Plants")


# would like to look at PERCENT plants alive in next analysis
# I also want to look at average biomass per plant per plot
# and average biomass per species per treatment


```


Continued analysis
```{r}

#calculating % plants alive in each plot

#filtering for only tarp treatment to see the total number of plants (both A and D)
tarp_filt <- elings_data %>% #calling on elings_data for the data frame
  filter(mortality_status == "D" | mortality_status == "A") %>% #filtering for non-NA values
  filter(treatment == "tarp") %>% #filtering for only tarped plants
  count(mortality_status) %>%
            bind_rows(summarise(., across(where(is.numeric), sum),
                                   across(where(is.character), ~'Total'))) %>% 
  rename(plant_count = n)


#filtering for only mulch treatment to see the total number of plants (both A and D)
mulch_filt <- elings_data %>% 
  filter(mortality_status == "D" | mortality_status == "A") %>% 
  filter(treatment == "mulch") %>% 
  count(mortality_status) %>%
            bind_rows(summarise(., across(where(is.numeric), sum),
                                   across(where(is.character), ~'Total'))) %>% 
  rename(plant_count = n)

# I want to find the % alive for tarped vs mulched plots -->tarp_perc = (mortality_status == "A" | "D")

```



## Future plans

Future analysis will include:

- looking at percent of plants alive in each plot and treatment

- analyzing average biomass per plant per plot

- average biomass per species per treatment

- compare san diego scrub species across treatments (bergaro, lycium, agave, amorpha, euphorbia) how are species that wouldn't normally be found at Elings

- maybe comparing species that are from further south, further inland, or tried and true SB natives

- compare mortality between provinces within treatments

- rates of growth in future surveys

- does decomposition correlate to plant growth in any way?


The analysis will (hopefully) answer these questions:

- Do tarped plots or mulched plots promote more growth?

- Which treatment has the lowest mortality rate?

- Which species performed the best (in terms of growth and survival) in each plot and each treatment? Is one treatment better for some plants, while the other treatment is better for other plants?

- Does origin impact plant mortality or growth?

- Which stpu location grew the most? Which stpu location had the best survival rates?
